Persistence in functional topology and a correction to a theorem of Morse

### 2021-09-01 sec/persistence.tex
Return to original paper size and use biblatex for references

Usual constructions that work in the category of vector spaces over the field $\F$ can be transferred to the category of persistence modules by applying them pointwise.
For example, the kernel and cokernel of a morphism, as well as the direct sum of persistence modules are well-defined.
Persistence modules that are \emph{indecomposable}, i.e., those that have only trivial direct sum decompositions, play an important role in persistence theory.
A rich family of indecomposable persistence modules is given by \emph{interval modules}, which for intervals $I \subseteq \R$ are defined by
\begin{equation} \label{e:interval module}
C(I)_t =
\begin{cases}
\F & \text{if } t \in I, \\
0 & \text{otherwise},
\end{cases}
\qquad \qquad
C(I)_{s, t} =
\begin{cases}
\operatorname{id}_{\F} & \text{if } s, t \in I, \\
0 & \text{otherwise}.
\end{cases}
\end{equation}
These indecomposable interval modules can be used as building blocks for \emph{barcode modules}, which are direct sums of interval modules.
Given a barcode module $\bigoplus_{\lambda \in \Lambda} C(I_{\lambda})$, the associated multiset of intervals $\{I_{\lambda}\}_{\lambda \in \Lambda}$ is known as its \emph{barcode}.
By a version of the Krull--Remak--Schmidt--Azumaya Theorem \cite{Azumaya.1950} (see also \cite[Theorem 2.7]{Chazal.2016a} for a specialization to barcode modules), two isomorphic barcode modules have the same barcodes up to a choice of the index set $\Lambda$.
Thus, if a persistence module $M$ is provided with an isomorphism to a barcode module, referred to as a \emph{barcode decomposition}, the associated barcode is a complete isomorphism invariant of $M$.
Hence, understanding which persistence modules admit barcode decompositions is key.

The most commonly used existence result for barcode decompositions is due to Crawley-Boevey's \cite{Crawley-Boevey.2015}.
It guarantees the existence of a barcode decomposition for any \emph{pointwise finite dimensional} ($\PFD$) persistence module, which is a persistence modules $M$ such that $M_t$ is a finite dimensional vector space for all $t \in \R$.
Unfortunately, the $\PFD$ condition is too restrictive for many purposes.
In particular, it is unsuited for the applications of Morse and Tompkins in minimal surface theory.
An appropriate weakening of $\PFD$ for more general settings is the notion of \emph{q-tameness}.
A persistence module $M$ is q-tame if the rank of the map $M_{s,t} \colon M_s \to M_t$ is finite for all $s < t$ \cite{Chazal.2016a}.
As exemplified by the infinite product of interval modules $\prod_{n \in \N_{> 0}} C([0,1/n))$, not every q-tame persistence module admits a barcode decomposition in the above sense.
Yet, there are multiple ways to regularize q-tame persistence modules to obtain discrete invariants similar to barcodes \cite{Chazal.2016a, Chazal.2016b, Schmahl.2021}.
We briefly recall the approach from \citet{Chazal.2016b}.

@@

\begin{thm}[\citet{Chazal.2016a, Chazal.2016b}] \label{t:q-tame modules have barcodes}
	Every q-tame persistence module has a unique persistence diagram that completely determines its isomorphism type in the observable category.
\end{thm}

>>>>

Usual constructions that work in the category of vector spaces over the field $\F$ can be transferred to the category of persistence modules by applying them pointwise.
For example, the kernel and cokernel of a morphism, as well as the direct sum of persistence modules are well-defined.
Persistence modules that are \emph{indecomposable}, i.e., those that have only trivial direct sum decompositions, play an important role in persistence theory.
A rich family of indecomposable persistence modules is given by \emph{interval modules}, which for intervals $I \subseteq \R$ are defined by
\begin{equation*}
C(I)_t =
\begin{cases}
\F & \text{if } t \in I, \\
0 & \text{otherwise},
\end{cases}
\qquad \qquad
C(I)_{s, t} =
\begin{cases}
\operatorname{id}_{\F} & \text{if } s, t \in I, \\
0 & \text{otherwise}.
\end{cases}
\end{equation*}
These indecomposable interval modules can be used as building blocks for \emph{barcode modules}, which are direct sums of interval modules.
Given a barcode module $\bigoplus_{\lambda \in \Lambda} C(I_{\lambda})$, the associated multiset of intervals $\{I_{\lambda}\}_{\lambda \in \Lambda}$ is known as its \emph{barcode}.
By a version of the Krull--Remak--Schmidt--Azumaya Theorem \cite{Azumaya.1950} (see also \cite[Theorem 2.7]{Chazal.2016a} for a specialization to barcode modules), two isomorphic barcode modules have the same barcodes up to a choice of the index set $\Lambda$.
Thus, if a persistence module $M$ is provided with an isomorphism to a barcode module, referred to as a \emph{barcode decomposition}, the associated barcode is a complete isomorphism invariant of $M$.
Hence, understanding which persistence modules admit barcode decompositions is key.

The most commonly used existence result for barcode decompositions is due to Crawley-Boevey's \cite{Crawley-Boevey.2015}.
It guarantees the existence of a barcode decomposition for any \emph{pointwise finite dimensional} ($\PFD$) persistence module, which is a persistence modules $M$ such that $M_t$ is a finite dimensional vector space for all $t \in \R$.
Unfortunately, the $\PFD$ condition is too restrictive for many purposes.
In particular, it is unsuited for the applications of Morse and Tompkins in minimal surface theory.
An appropriate weakening of $\PFD$ for more general settings is the notion of \emph{q-tameness}.
A persistence module $M$ is q-tame if the rank of the map $M_{s,t} \colon M_s \to M_t$ is finite for all $s < t$ \cite{Chazal.2016a}.
As exemplified by the infinite product of interval modules $\prod_{n \in \N_{> 0}} C([0,1/n))$, not every q-tame persistence module admits a barcode decomposition in the above sense.
Yet, there are multiple ways to regularize q-tame persistence modules to obtain discrete invariants similar to barcodes \cite{Chazal.2016a, Chazal.2016b, Schmahl.2021}.
We briefly recall the approach from \cite{Chazal.2016b}.

@@

\begin{thm}[\cite{Chazal.2016a, Chazal.2016b}] \label{t:q-tame modules have barcodes}
	Every q-tame persistence module has a unique persistence diagram that completely determines its isomorphism type in the observable category.
\end{thm}

### 2021-09-01 sec/surfaces.tex
Return to original paper size and use biblatex for references

\begin{figure}[t]
	\centering
	\begin{tikzpicture}[scale = 76]
	\draw[thick] (-.1,-.001) rectangle (.1,.05);
	\clip (-.1,0) rectangle (.1,.05);
	\foreach \i in {1,...,100}{
		\draw[line width=0.1/\i mm] (0, 1/\i^2) circle (1/\i^2);
	}
	\end{tikzpicture}
	\caption{A closeup of the Hawaiian earring $\mathbb{H}^1$.}
\end{figure}

@@

	It remains to be shown that $f_{\leq \bullet}$ is not q-tame for $\H$.
	This follows directly from our assumption that $\H_{n}(\HE)$ is not finite dimensional for some $n$, as $f_{\leq t}$ is constant with value $\HE$ for $t \geq 1$.
\end{proof}
\newpage
Both singular and \v{C}ech homology of the $d$-dimensional Hawaiian earring is infinite dimensional in degree $d$, so they satisfy the assumptions of \cref{t:counterexample}.
For singular homology this is proven in \cite{Barratt.1962}.
For \v{C}ech homology, one can use the fact that it commutes with totally ordered limits for compact Hausdorff spaces \cite[Theorems VIII.3.6.\@ and X.3.1.]{Eilenberg.1952} as follows.
Define 
\begin{align*}
\HE_k &=
\left\{ (x_0, \dots, x_d) \in \R^{d+1} \ \middle| \ \left( x_0 - \frac{1}{k} \right)^2 + x_1^2 + \dots + x_d^2 \leq \left( \frac{1}{k} \right)^2 \right\} \\ &\, \cup
\bigcup_{n=1}^{k-1} \left\{ (x_0, \dots, x_d) \in \R^{d+1} \ \middle |\ \left( x_0 - \frac{1}{n} \right)^2 + x_1^2 + \dots + x_d^2 = \left( \frac{1}{n} \right)^2 \right\},
\end{align*}
i.e., the $d$-dimensional Hawaiian earring but with the $k$-th largest $d$-sphere filled.
We have $\lim_{k} \HE_{k} = \bigcap_{k} \HE_{k} = \HE$, and hence $\CH_{d}(\HE; \F) = \lim_{k} \CH_{d}(\HE_{k}; \F)$, where $\CH$ denotes \v{C}ech homology.
Clearly, each $\HE_{k}$ is a CW-complex, so we can simply use cellular homology to compute
\begin{equation*}
\lim_{k}\CH_{d}(\mathbb{H}^{d}_{k}; \F)=\lim\left(\dots\to \prod_{n=1}^2\F\to \prod_{n=1}^1\F\to \prod_{n=1}^0\F\right)=\prod_{n\in\N}\F,
\end{equation*}
which is infinite-dimensional over $\F$.

>>>>

\begin{figure}[t]
	\centering
	\begin{tikzpicture}[scale = 60]
	\draw[thick] (-.1,-.001) rectangle (.1,.05);
	\clip (-.1,0) rectangle (.1,.05);
	\foreach \i in {1,...,100}{
		\draw[line width=0.1/\i mm] (0, 1/\i^2) circle (1/\i^2);
	}
	\end{tikzpicture}
	\caption{A closeup of the Hawaiian earring $\mathbb{H}^1$.}
\end{figure}

@@

	It remains to be shown that $f_{\leq \bullet}$ is not q-tame for $\H$.
	This follows directly from our assumption that $\H_{n}(\HE)$ is not finite dimensional for some $n$, as $f_{\leq t}$ is constant with value $\HE$ for $t \geq 1$.
\end{proof}

Both singular and \v{C}ech homology of the $d$-dimensional Hawaiian earring is infinite dimensional in degree $d$, so they satisfy the assumptions of \cref{t:counterexample}.
For singular homology this is proven in \cite{Barratt.1962}.
For \v{C}ech homology, one can use the fact that it commutes with totally ordered limits for compact Hausdorff spaces \cite[Theorems VIII.3.6.\@ and X.3.1.]{Eilenberg.1952} as follows.
Define
\begin{align*}
\HE_k &=
\left\{ (x_0, \dots, x_d) \in \R^{d+1} \ \middle| \ \left( x_0 - \frac{1}{k} \right)^2 + x_1^2 + \dots + x_d^2 \leq \left( \frac{1}{k} \right)^2 \right\} \\ &\, \cup
\bigcup_{n=1}^{k-1} \left\{ (x_0, \dots, x_d) \in \R^{d+1} \ \middle |\ \left( x_0 - \frac{1}{n} \right)^2 + x_1^2 + \dots + x_d^2 = \left( \frac{1}{n} \right)^2 \right\},
\end{align*}
i.e., the $d$-dimensional Hawaiian earring but with the $k$-th largest $d$-sphere filled.
We have $\lim_{k} \HE_{k} = \bigcap_{k} \HE_{k} = \HE$, and hence $\CH_{d}(\HE; \F) = \lim_{k} \CH_{d}(\HE_{k}; \F)$, where $\CH$ denotes \v{C}ech homology.
Clearly, each $\HE_{k}$ is a CW-complex, so we can simply use cellular homology to compute
\begin{equation*}
\lim_{k}\CH_{d}(\mathbb{H}^{d}_{k}; \F)=\lim\left(\dots\to \prod_{n=1}^2\F\to \prod_{n=1}^1\F\to \prod_{n=1}^0\F\right)=\prod_{n\in\N}\F,
\end{equation*}
which is infinite-dimensional over $\F$.
